Astrophysical calibration of gravitational-wave detectors

### 2015-07-28 docs/cosmic_cal.tex
cosmic_cal.tex: added comment
Some minor updates to the paper

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 1000\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity (as given by Figure 1 of \cite{2013arXiv1304.0670L}). As such we add noise to each 
simulation based on the sensitivity curve, but scaled with the associated calibration scale factor.

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed 2000 simulations at each 
distance value. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, and with the scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on it having an SNR of $\geq 5.5$ 
in at least two detectors as used in \cite{2012PhRvD..85h2002A} (as these would be signals 
coincident with a GRB we do not include the often used further constraint that the total coherent 
network SNR is greater than 12). The coalescence times for all injections were held fixed 
within the centre of the prior window. This criteria means that as we go to larger distances our 
population of injected sources will not be uniformly distributed over the sky, and will generally 
be closer to being circularly polarised. For the \ac{NSBH} systems it also means that at larger 
distances we will have more sources with larger mass ratios \MP{Check this}.

@@

A simple assumption that one could make would be that the uncertainties on the calibration scale 
factors (if they are independent of other parameters and have a Gaussian probability distribtuion) 
should be given by $\sim 1/{\rm SNR}$ for each detector. As we will show below this assumption is 
reasonable, but correlations do exist between parameters meaning that it does not completely hold, 
especially in the case on \ac{NSBH} systems with a spinning component.

For each of the simulated sources we have used the {\tt emcee} python MCMC package to perform 
parameter estimation over the unknown source parameters using the prior as discussed in 
Section~\ref{sec:priors}. In each case we use an estimate of the noise PSD based on the design 
sensitivity, but simulated as if estimated by averaging 64 separate PSD estimates and scaled with 
the same factor as applied to the injection and noise\footnote{We do not account for there being 
a difference between the estimated PSD and the actual PSD of the analysed section of data. This 
difference would be very much correlated with the scale factor, so in reality our estimate of the 
scale factor would be a combination of the calibration offset and the difference in the PSD. As 
such our results on real data would be an upper limit on the calibration scale factor.}. This has 
provided posterior probability distributions on the calibration scale factors for each detector 
\MP{Show an example for both source types}. Examples of the marginalised posterior probabilities
for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network are shown in 
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the scale factors 
for each detector (if these were Gaussian distributions this is equivalent to the region bounded by 
$1\sigma$ intervals). For all the signals at each distance increment we have 
produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of these scale factors 
confidence intervals compared to the actual value. These are shown as boxplots in 
figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent the range of values. The black line within each box gives the median value and 
the star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage 
of proposed injections that actually fulfilled the detection criterion.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{bns_post_fig.pdf}
 \end{center}
 \caption{\label{fig:bnspost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{BNS} system including calibration scaling factors for the three detectors (H1, 
L1 and V1). The injected signal was at a distance of 250\,Mpc, had \ac{SNR} of 8.7, 10.9 and 4.1 
and percentage uncertainties in the calibration scale factors of 14\%, 11\% and 27\% for each of 
the detectors respectively.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{nsbh_post_fig.pdf}
 \end{center}
 \caption{\label{fig:nsbhpost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{NSBH} system including calibration scaling factors for the three detectors 
(H1, L1 and V1). The injected signal was at a distance of 450\,Mpc, had \ac{SNR} of 7.7, 10.9 and 
4.2 and percentage uncertainties in the calibration scale factors of 17\%, 14\% and 34\% for each of 
the detectors respectively.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factor can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The boxplots span the lower to 
upper quartile range of the distributions, with the median value shown as a horizontal line within 
the box and the mean shown as a star. The dashed magenta line shows the percentage of source drawn 
from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factor can be determined for a three detector network if assuming \ac{NSBH} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The plot contents are the same as 
in figure~\ref{fig:bnsresults}.}
\end{figure}

@@

In figure~\ref{fig:nsbhresults} we see slightly different results. Despite the generally higher SNR 
for the \ac{NSBH} systems we find that on averge that the scale factor uncertainty is still roughly 
20\% at distances to 400--500\,Mpc. The range of uncertainty values is also far broader at all 
distances and there is not the plateau and upper limit seen for the \ac{BNS} systems. \MP{This 
needs checked, but ... This broader range of uncertainties appears to be down to the inclusion of 
the spin parameter and correlations this has with the mass ratio $q$ - show posterior examples. The 
lack of an upper plateau seems to be due to the detection criteria having a different effect on the 
distribution of sources that could be seen, in particular it biases the observable sources towards 
having a higher mass ratio (again check this) and thus having more correlation with the spin.}

>>>>

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 900\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity (as given by Figure 1 of \cite{2013arXiv1304.0670L}). As such we add noise to each 
simulation based on the sensitivity curve, but scaled with the associated calibration scale factor.

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed up to 2000 simulations at each 
distance value\footnote{Simulations on which our MCMC did not produce converged chains within our 
analysis time were discarded and not re-run, although there were always at least 1000 simulations 
that completed.}. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, and with the scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on it having an SNR of $\geq 5.5$ 
in at least two detectors as used in \cite{2012PhRvD..85h2002A} (as these would be signals 
coincident with a GRB we do not include the often used further constraint that the total coherent 
network SNR is greater than 12). The coalescence times for all injections were held fixed 
within the centre of the prior window. This criteria means that as we go to larger distances our 
population of injected sources will not be uniformly distributed over the sky, and will generally 
be closer to being circularly polarised. For the \ac{NSBH} systems it also means that at larger 
distances we will have more sources with larger mass ratios \MP{Check this}.

@@

A simple assumption that one could make would be that the uncertainties on the calibration scale 
factors (if they are independent of other parameters and have a roughly Gaussian probability 
distribtuion)
should be given by $\sim 1/{\rm SNR}$ for each detector. As we will show below this assumption is 
reasonable, but small correlations do exist between parameters meaning that it does not completely 
hold.

For each of the simulated sources we have used the {\tt emcee} python MCMC package 
\cite{2013PASP..125..306F} to perform parameter estimation over the unknown source parameters using 
the priors as discussed in Section~\ref{sec:priors}. In each case when calculating the likelihood 
we used an estimate of the noise \ac{PSD} based on the advanced detector design sensitivities 
(using those given in \cite{2013arXiv1304.0670L}), but simulated as if calculated by averaging 64 
separate noisy \ac{PSD} estimates and scaled with the same factor as applied to the injection and 
noise\footnote{We do not account for there being a difference between the estimated \ac{PSD} and the 
actual \ac{PSD} of the analysed section of data. This difference would be very much correlated with 
the scale factor, so in reality our estimate of the scale factor would be a combination of the 
calibration offset and the difference in the \ac{PSD}. As such our results on real data would be an 
upper limit on the calibration scale factor.}. This has provided posterior probability distributions 
on the calibration scale factors for each detector. Examples of the marginalised posterior 
probabilities for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network 
are shown in figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the 
calibration scale factors for each detector (if these were Gaussian distributions this is 
equivalent to the region bounded by $1\sigma$ intervals). For all the signals at each distance 
increment we have produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of 
these scale factor's confidence intervals compared to the actual value. These are shown as boxplots 
in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent the range of values. The black line within each box gives the median value and 
the star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage 
of signal drawn from the prior distribution that would fulfill the detection criterion.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{bns_post_fig.pdf}
 \end{center}
 \caption{\label{fig:bnspost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{BNS} system, including calibration scaling factors for the three detectors 
(H1, L1 and V1). The injected signal was at a distance of 250\,Mpc, had \ac{SNR} of 8.7, 10.9 and 
4.1 and percentage uncertainties in the calibration scale factors of 14\%, 11\% and 27\% for each 
of the detectors respectively.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{nsbh_post_fig.pdf}
 \end{center}
 \caption{\label{fig:nsbhpost} The marginalised posterior probability distributions for the unknown
 parameters of a \ac{NSBH} system, including calibration scaling factors for the three detectors 
(H1, L1 and V1). The injected signal was at a distance of 450\,Mpc, had \ac{SNR} of 7.7, 10.9 and 
4.2 and percentage uncertainties in the calibration scale factors of 17\%, 14\% and 34\% for each of 
the detectors respectively.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The boxplots span the lower to 
upper quartile range of the distributions, with the median value shown as a horizontal line within 
the box and the mean shown as a star. The dashed magenta line shows the percentage of source drawn 
from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{NSBH} systems 
(provided a coincident GRB is observed and can yield a distance estimate). The plot contents are 
the same as 
in figure~\ref{fig:bnsresults}.}
\end{figure}

@@

In figure~\ref{fig:nsbhresults} we see very similar results although the higher \ac{SNR} of the 
signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on the calibration scale 
factors for H1 and L1 out to greater distances.
